Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan

### 2025-03-10 analysis/paper/paper.qmd
finalizing the paper

---
title: "Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Institute of History and Philology, Academia Sinica
title-block-published: "Last updated"
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Colonial encounters between Europeans and Indigenous peoples often had profound impacts on Indigenous societies and cultures in many parts of the world. In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foriegn contact through regional trade networks in East Asia, which expanded further with the arrival of Europeans in the 17th century. This study focuses on local subsistence through faunal remains to explore how Indigenous communities responded to trade activities driven by Europeans and Han Chinese, particularly the high demand for natural resources such as deer. By analyzing faunal remains from the upper component of Kiwulan (AD 1350-1850), a late Iron Age site in northeastern Taiwan, this research dicusses the Indigenous participation in the deerskin trade and their adaptive strategies. The results suggest cultural persistence in deer consumption, along with increased reliance on suids during the European period as demand for deer products intensified. This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European colonial territories, a perspective rarely explored.

@@

```{r}
#| label: taxa-NISP-counts
#| echo: false
# taxa counts
fauna_combined_taxa <-
  fauna_combined_context_all %>%
  count(taxa, `Weight (g)`) %>%
  mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
                          TRUE ~ taxa)) %>%
  mutate(`Common name` = case_when(str_detect(taxa, "Cer") ~ "Sika deer", str_detect(taxa, "Bub") ~ 
                                   "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",
                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Suid",
                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~ 
                                   "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~ 
                                   "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
                                   str_detect(taxa, "cervid") ~ "Deer", taxa == "Sus scrofa domestica" ~ 
                                   "Domestic pig", taxa == "Aves" ~ "Birds")) %>%
  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
                           str_detect(taxa, "Sus") ~ 3, str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 4,
                           str_detect(taxa, "munt") ~ 5, str_detect(taxa, "Rat") ~ 6, 
                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Tes") ~ 8,
                           str_detect(taxa, "Ave") ~ 9, str_detect(taxa, "large m") ~ 10,
                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
  group_by(group) %>%
  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
  ungroup () %>%
  rename(Taxon = taxa, NISP = n) %>%
  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
  bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>% 
  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`)) %>% 
  mutate(Taxon = ifelse(Taxon == "Sus scrofa", "Sus spp.", Taxon))

total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "In")) %>% summarise(sum(NISP))
deer_NISP <- fauna_combined_taxa %>% 
  filter(str_detect(`Common name`, "eer")) %>% summarise(sum(NISP)) #|str_detect(Taxon, "Ree")
suid_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Sus spp.")) %>% summarise(sum(NISP))
buff_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Bub")) %>% summarise(sum(NISP))
deer_NISP_per <- round(deer_NISP/(total_NISP-unident), 2)*100
suid_NISP_per <- round(suid_NISP/(total_NISP-unident), 2)*100
buff_NISP_per <- round(buff_NISP/(total_NISP-unident), 2)*100
```

The faunal assemblage analyzed in this study from Kiwulan consists of `r total_NISP` bone fragments based on NISP (@tbl-counts). Among the identifiable bones, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`) and water buffalo (`r buff_NISP_per`). To further compare general patterns of animal consumption in terms of their economic values over time, animals within the same family with of similar body size were grouped together. Therefore, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were grouped into the bovine category. Similarly, domestic pig (n = 2, from CL4 and CL6) were integrated into the suid category. Rats (Rattus sp.) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes. In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 ( (@tbl-counts, @fig-NISP-proportion: A). A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).

@@

<!-- population dynamics and arguments-->
The high population during the European period, as inferred by the frequencies of pottery, may have contributed to the overall increase in animal use [@Wang2020pot; @Hsieh2009]. However, the impact of foreign presence can be still be identified in deer butchery patterns, the relative proportions of animals, and animal ubiquity. First, lower limbs with skinning cuts were mostly found in the European period, followed by the late pre-European phase, aligning with the rise of regional trade in the 14th–15th centuries and its intensification in the 17th century. Second, if the high frequency of deer heads during the European period was solely due to food stress from population growth, leading to increased brain or tongue consumption, a parallel rise in marrow extraction would be expected. However, long bones with fracture patterns showing 10–30% less frequent during the European and late pre-European periods than in other periods. Third, population growth would have likely led to a more proportional increase in both large and small animals. Despite high taxonomic diversity during the European period, suids showed a disproportionate rise in ubiquity, quantity, and proportion. This indicates that the trade of deerskin and meat may have led Indigenous people to rely more on suids as an alternative prey due to their comparable body size and meat return rates to deer [@Broughton2011prey]. The role of suids as a secondary resource after deer has been observed in Indigenous cultures in western Taiwan [@Chuang2021], suggesting they may have served as a supplement when deer were scarce or limited, possibly because some deer were reserved for trade.

>>>>

---
title: "Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Colonial encounters between Europeans and Indigenous peoples often had profound impacts on Indigenous societies and cultures in many parts of the world. In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foriegn contact through regional trade networks in East Asia, which expanded further with the arrival of Europeans in the 17th century. This study focuses on local subsistence through faunal remains to explore how Indigenous communities responded to trade activities driven by Europeans and Han Chinese, particularly the high demand for natural resources such as deer. By analyzing faunal remains from the upper component of Kiwulan (AD 1350-1850), a late Iron Age site in northeastern Taiwan, this research dicusses the Indigenous participation in the deerskin trade and their adaptive strategies. The results suggest cultural persistence in deer consumption, along with increased reliance on suids during the European period as demand for deer products intensified. This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European colonial territories, a perspective rarely explored.

@@

```{r}
#| label: taxa-NISP-counts
#| echo: false
# taxa counts
fauna_combined_taxa <-
  fauna_combined_context_all %>%
  count(taxa, `Weight (g)`) %>%
  mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
                          TRUE ~ taxa)) %>%
  mutate(`Common name` = case_when(str_detect(taxa, "Cer") ~ "Sika deer", str_detect(taxa, "Bub") ~
                                     "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",
                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Suid",
                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~
                                     "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~
                                     "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
                                   str_detect(taxa, "cervid") ~ "Deer", taxa == "Sus scrofa domestica" ~
                                     "Domestic pig", taxa == "Aves" ~ "Birds")) %>%
  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
                           str_detect(taxa, "Sus") ~ 3, str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 4,
                           str_detect(taxa, "munt") ~ 5, str_detect(taxa, "Rat") ~ 6,
                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Tes") ~ 8,
                           str_detect(taxa, "Ave") ~ 9, str_detect(taxa, "large m") ~ 10,
                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
  group_by(group) %>%
  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
  ungroup () %>%
  rename(Taxon = taxa, NISP = n) %>%
  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
  bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>%
  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`)) %>%
  mutate(Taxon = ifelse(Taxon == "Reeves's muntjac", "Muntiacus reevesi", Taxon)) %>%
  mutate(Taxon = ifelse(Taxon == "Cervus nippon", "Cervus nippon taiouanus", Taxon))

total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "In")) %>% summarise(sum(NISP))
deer_NISP <- fauna_combined_taxa %>% 
  filter(str_detect(`Common name`, "eer")) %>% summarise(sum(NISP)) #|str_detect(Taxon, "Ree")
suid_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Sus")) %>% summarise(sum(NISP))
buff_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Bub")) %>% summarise(sum(NISP))
deer_NISP_per <- round(deer_NISP/(total_NISP-unident), 2)*100
suid_NISP_per <- round(suid_NISP/(total_NISP-unident), 2)*100

buff_NISP_per <- round(buff_NISP/(total_NISP-unident), 2)*100
```

The faunal assemblage analyzed in this study from Kiwulan consists of `r total_NISP` bone fragments based on NISP (@tbl-counts). Among the identifiable bones, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`%) and water buffalo (`r buff_NISP_per`%). To further compare general patterns of animal consumption in terms of their economic values over time, animals within the same family with of similar body size were grouped together. Therefore, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were grouped into the bovine category. Similarly, domestic pigs (CL4: n = 2, CL6: n = 1) were integrated into the suid category. Rats (Rattus sp.) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes. In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 ( (@tbl-counts, @fig-NISP-proportion: A). A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).

@@

<!-- population dynamics and arguments-->
The high population during the European period, as inferred by the frequencies of pottery, may have contributed to the overall increase in animal use [@Wang2020pot; @Hsieh2009]. However, the impact of foreign presence can be still be identified in deer butchery patterns, the relative proportions of animals, and animal ubiquity. First, lower limbs with skinning cuts were mostly found in the European period, followed by the late pre-European phase, aligning with the rise of regional trade in the 14th–15th centuries and its intensification in the 17th century. Second, if the high frequency of deer heads during the European period was solely due to food stress from population growth, leading to increased brain or tongue consumption, a parallel rise in marrow extraction would be expected. However, long bones with fracture patterns showing 10–30% less frequent during the European and late pre-European periods than in other periods. Third, population growth would have likely led to a more proportional increase in both large and small animals. Despite high taxonomic diversity during the European period, suid remains showed a disproportionate rise in ubiquity, quantity, and proportion. This indicates that the trade of deerskin and meat may have led Indigenous people to rely more on suids as an alternative prey due to their comparable body size and meat return rates to deer [@Broughton2011prey].

The role of suids as a secondary resource after deer has been observed in the Indigenous cultures in southwestern Taiwan during the Neolithic period and Iron Age [@Chuang2021], suggesting that suids may have served as a supplementary resource while deer remained the prioritized prey. At Kiwulan, it is possible that portions of deer were reserved for trade, limiting local consumption and leading to an increased reliance on suids. Domestic pigs were identified at Kiwulan in both the European and pre-European periods based on their much smaller third molar size (length < 30 mm) and morphology, in comparison to archaeological data from Neolithic and Iron Age sites in Taiwan [@KtLi2015; @Chuang2021]. However, a clear distinction between wild boars and domestic pigs in archaeological remains is challenging due to overlapping size ranges, regional variations, and the potential presence of hybrids [@Rowley2012; @Chuang2021]. The third molar lengths of suids found at Kiwulan (28–36 mm) may reflect such hybridization patterns. While current evidence does not clearly indicate the emergence of intensified suid management, an increase in suid utilization, either through domestication or hunting of wild individuals, is evident during the European period.
